Continuous risk-based predictive approaches to treatment effect heterogeneity: A simulation study

### 2021-06-22 submission/manuscript.rmd
add corrections to results section

When we increased the sample size (N = `r 17000`), the model including a
constant relative treatment effect had the lowest RMSE under the base case
scenario of constant relative treatment effects (Figure
\ref{fig:rmsesamplesize}; Panel A). When introducing moderate and strong linear
deviations the linear interaction model performed best (Figure
\ref{fig:rmsesamplesize}; Panels B and C). In the case of true strong quadratic
deviations from the base case of constant relative effects, the more flexible
RCS smoothing models (3 knots) had the lowest RMSE. Again, the increased
flexibility of RCS smoothing with higher number of knots resulted in overfitting
and worse performance (Figure \\ref{fig:rmsesamplesize}; Panel D).

@@

All candidate methods demonstrated comparable discrimination for benefit in all
scenarios where linear and quadratic deviations from the base case of constant
treatment effect were considered (N=`r 4250`; AUC = `r 0.75`). However, models
including a linear interaction with the risk linear predictor tended to present
much lower variability compared to all other model-based and smoothing
approaches (Figure \ref{fig:discrimination}). We also observed an increasing
trend of discrimination for benefit variability with increasing number of
restricted cubic spline knots in all scenarios.

@@

When focusing on calibration, the linear interaction model had the lowest median
ICI for benefit in the majority of scenarios except for the scenarios where
moderate linear deviations from the base case were considered. In that case
constant treatment effect models demonstrated the best performance, very
comparable to the linear interaction model's performance, nonetheless (Figure
\ref{fig:calibration}; Panels A, B, and C).

@@

## Real data
```{r, echo=FALSE, warning=FALSE, message=FALSE}
load(here::here("data/raw/gusto.rda"))
gusto <- gusto %>%
  tibble() %>%
  filter(!is.na(tpa))

@@

# Discussion
We have carried out a simulation study comparing different approaches for
individualized benefit prediction using baseline risk as the reference. These
methods included a constant treatment effect model, a model including a simple
linear interaction of the risk linear predictor with treatment, and models using
smoothing with restricted cubic splines with 3, 4, and 5 knots. Performance was
assessed evaluating error rates using RMSE and benefit prediction performance
using some novel methods for evaluating discrimination and calibration for
benefit.

The linear interaction model proved to be flexible enough to adequately predict
benefit under a wide range of scenarios. It had lower error rates and usually
the highest c-for-benefit with the lower variability across simulation
replications. Its benefit predictions were also well-calibrated across the
majority of the simulation settings. Models with RCS smoothing with 3 knots
performed very similarly, only outperformig the linear interaction model in the
presence of strong quadratic deviations from the base case scenario of constant
relative treatment effects. These simpler methods proved to have quite robust
performance across all scenarios, which was not the case for more flexible
approaches.

@@

In conclusion, we carried out a simulation study comparing different risk-based
approaches to individualizing benefit predictions. Simpler models including a
linear interaction with the risk linear predictor or RCS smoothing with 3 knots
had adequate performance under a wide range scenarios. An adaptive approach based
on selecting among the considered methods using AIC had satisfactory
performance.

>>>>

When we increased the sample size (N = `r 17000`), the model including a
constant relative treatment effect had the lowest RMSE under the assumption of
true constant relative treatment effects (Figure \ref{fig:rmsesamplesize};
Panel A). Contrary to previous results with smaller sample sizes, when
introducing moderate and strong linear deviations the linear interaction model
performed best, outperforming the constant treatment effect model (Figure
\ref{fig:rmsesamplesize}; Panels B and C). However, the linear interaction
model was outperformed by the more flexible RCS models (3 knots) in the case of
strong quadratic deviations. Again, the increased flexibility of RCS smoothing
with higher number of knots resulted in overfitting and worse performance
(Figure \ref{fig:rmsesamplesize}; Panel D).

@@

The constant effects model, the linear interaction model and models with RCS
smoothing (3 knots) had the highest median c-for-benefit in the base case
scenario and the scenarios where linear and quadratic deviations were
considered. However, constant treatment effects and linear interaction models
tended to present much lower variability compared to all other model-based and
smoothing approaches (Figure \ref{fig:discrimination}). We also observed an
increasing trend of discrimination for benefit variability and decreasing median
values with increasing number of restricted cubic spline knots in all scenarios.

@@

When focusing on calibration, the linear interaction model had the lowest median
ICI for benefit in the majority of scenarios except for the scenarios where no
or moderate linear deviations from the base case were considered. In that case
constant treatment effect models demonstrated the best performance, very
comparable to the linear interaction model's performance, nonetheless (Figure
\ref{fig:calibration}). However, under strong linear or
quadratic deviations, the constant treatment effect model was very poorly
calibrated (Figure \ref{fig:calibration}; Panels C and D).

@@

## Case study
```{r, echo=FALSE, warning=FALSE, message=FALSE}
load(here::here("data/raw/gusto.rda"))
gusto <- gusto %>%
  tibble() %>%
  filter(!is.na(tpa))

@@

# Discussion

The linear interaction model proved to be flexible enough to adequately predict
benefit under a wide range of scenarios. It had lower error rates and usually
the highest c-for-benefit with the lower variability across simulation
replications. Its benefit predictions were also well-calibrated across the
majority of the simulation settings. Models with RCS smoothing with 3 knots
performed very similarly, only outperforming the linear interaction model in the
presence of strong quadratic deviations from the base case scenario of constant
relative treatment effects. These simpler methods proved to have quite robust
performance across all scenarios, which was not the case for more flexible
approaches.

@@

In conclusion, when comparing different risk-based approaches to individualizing
benefit predictions simpler models including a linear interaction with the
prognostic index or RCS smoothing with 3 knots had adequate performance under a
wide range scenarios. An adaptive approach selecting among the considered
methods with optimal AIC had satisfactory performance.

### 2021-06-22 submission/manuscript.tex
add corrections to results section

When we increased the sample size (N = 17,000), the model including a
constant relative treatment effect had the lowest RMSE under the base
case scenario of constant relative treatment effects (Figure
\ref{fig:rmsesamplesize}; Panel A). When introducing moderate and strong
linear deviations the linear interaction model performed best (Figure
\ref{fig:rmsesamplesize}; Panels B and C). In the case of true strong
quadratic deviations from the base case of constant relative effects,
the more flexible RCS smoothing models (3 knots) had the lowest RMSE.
Again, the increased flexibility of RCS smoothing with higher number of
knots resulted in overfitting and worse performance (Figure
\textbackslash ref\{fig:rmsesamplesize\}; Panel D).

@@

All candidate methods demonstrated comparable discrimination for benefit
in all scenarios where linear and quadratic deviations from the base
case of constant treatment effect were considered (N=4,250; AUC = 0.75).
However, models including a linear interaction with the risk linear
predictor tended to present much lower variability compared to all other
model-based and smoothing approaches (Figure \ref{fig:discrimination}).
We also observed an increasing trend of discrimination for benefit
variability with increasing number of restricted cubic spline knots in
all scenarios.

@@

\hypertarget{real-data}{%
\subsection{Real data}\label{real-data}}

@@

We have carried out a simulation study comparing different approaches
for individualized benefit prediction using baseline risk as the
reference. These methods included a constant treatment effect model, a
model including a simple linear interaction of the risk linear predictor
with treatment, and models using smoothing with restricted cubic splines
with 3, 4, and 5 knots. Performance was assessed evaluating error rates
using RMSE and benefit prediction performance using some novel methods
for evaluating discrimination and calibration for benefit.

The linear interaction model proved to be flexible enough to adequately
predict benefit under a wide range of scenarios. It had lower error
rates and usually the highest c-for-benefit with the lower variability
across simulation replications. Its benefit predictions were also
well-calibrated across the majority of the simulation settings. Models
with RCS smoothing with 3 knots performed very similarly, only
outperformig the linear interaction model in the presence of strong
quadratic deviations from the base case scenario of constant relative
treatment effects. These simpler methods proved to have quite robust
performance across all scenarios, which was not the case for more
flexible approaches.

>>>>

When we increased the sample size (N = 17,000), the model including a
constant relative treatment effect had the lowest RMSE under the
assumption of true constant relative treatment effects (Figure
\ref{fig:rmsesamplesize}; Panel A). Contrary to previous results with
smaller sample sizes, when introducing moderate and strong linear
deviations the linear interaction model performed best, outperforming
the constant treatment effect model (Figure \ref{fig:rmsesamplesize};
Panels B and C). However, the linear interaction model was outperformed
by the more flexible RCS models (3 knots) in the case of strong
quadratic deviations. Again, the increased flexibility of RCS smoothing
with higher number of knots resulted in overfitting and worse
performance (Figure \ref{fig:rmsesamplesize}; Panel D).

@@

The constant effects model, the linear interaction model and models with
RCS smoothing (3 knots) had the highest median c-for-benefit in the base
case scenario and the scenarios where linear and quadratic deviations
were considered. However, constant treatment effects and linear
interaction models tended to present much lower variability compared to
all other model-based and smoothing approaches (Figure
\ref{fig:discrimination}). We also observed an increasing trend of
discrimination for benefit variability and decreasing median values with
increasing number of restricted cubic spline knots in all scenarios.

@@

\hypertarget{case-study}{%
\subsection{Case study}\label{case-study}}

@@

The linear interaction model proved to be flexible enough to adequately
predict benefit under a wide range of scenarios. It had lower error
rates and usually the highest c-for-benefit with the lower variability
across simulation replications. Its benefit predictions were also
well-calibrated across the majority of the simulation settings. Models
with RCS smoothing with 3 knots performed very similarly, only
outperforming the linear interaction model in the presence of strong
quadratic deviations from the base case scenario of constant relative
treatment effects. These simpler methods proved to have quite robust
performance across all scenarios, which was not the case for more
flexible approaches.
